Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-18 manuscript.tex
Updates from Overleaf

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides, Roman o2013DNADependence, Hinckley2014Coarse-grainedEffects}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
% this section should look more like a conclusion foreshadowing conclusions and main take aways 
We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVS as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.

@@

% need some more detail here -- talk to Andy about this
MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. After Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.

@@

As an initial validation of the 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}. We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We derived relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We used this temperature shift to inform our equilibrium simulations in the proceeding section. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits. At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

Following the analysis pipeline described above, we generated SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the conditional transition probability to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

%The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}. The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails. Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to macrostate results, however, the 3S2 state is more stable than 5S2 states. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

% Although NN calculations predict 5' vs. 3' stabilities for AT-all, we observe that GC-end 5S2 state remains more populated than the 3S2 state contrary to NN predictions. Indeed, we do not observe the formation of a stable 3S4 state for GC-end, further indicating that these states 3' states are unfavorable. This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

%if AT-all has highest frequency and GC-end has lowest then can describe role of metastable states as increasing rate for the former but decreasing rate for the later (for kinetic traps that do not terminate in native hybridization and delay the process as a whole).

@@

The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D. Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed. On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation. We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently then in experiment. Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.

% discuss computational comparisons needs work
Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA}. Araque et al. defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}. When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al. found that the GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}. As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core. While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs \citep{Phys2019}. Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}. We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.

@@

To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near MSM-identified hybridization and dissociation events. We present two representative events in Figure \ref{fig:GC-mix_transitions} along with all native base pair distances over time. During hybridization, we observed the formations of some key base pair contacts before the full duplex formed. Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}. For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

@@

To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states. To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event. We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences. We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance. Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence. This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.

>>>>

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

We find that 3spn2 enables access to long timescales in which we observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence -- and therefore the kinetic accessibility and thermodynamic stability of non-native states -- leads to richer dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA studies, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.

@@

% need some more detail here -- talk to Andy about this
MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.

@@

As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}. We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

We followed the SRV-MSM analysis pipeline described above to generate sequence-dependent kinetic models for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the probability of a transition to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

%The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails. Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}.

% This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D. Thus once a D $\rightarrow$ F4 transition has occurred, a F4 $\rightarrow$ H transition will likely proceed. On the other hand, H $\rightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation. We previously noted that frayed dynamics are substantially accelerated in the 3spn2 model, which may lead to the F4 state being sampled more frequently than in experiment. Because GC-core stability is most sensitive to fraying, this effect could cause GC-core to dissociate more frequently and lead to the deviation in slow response we note in figure 2.

% discuss computational comparisons needs work
Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019,Sanstead2020OxidizedDNA}. Araque et al. defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}. When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al. found that GC-core had the highest deviation from two-state behavior during dissociation (when excluding out-of-register contributions) \citep{Sanstead2016}. As T-jump IR data only showed a loss of A:T contacts during the fast response, this intermediate state was defined by a high degree of fraying about the central core. While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, their lattice model predicted GC-core to have substantially more frayed base pairs \citep{Phys2019}. Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}. We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however our SRV-MSM results show this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.

@@

To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near MSM-identified hybridization and dissociation events. We present two representative events in Figure \ref{fig:GC-mix_transitions} along with all native base pair distances over time. During hybridization, we observed the some key base pair contacts before the full duplex formed. Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Sambriski2009SequencePathways,  Yin2011KineticsHybridization}. For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

@@

To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$H for GC-core) events and excluded transitions that passed through long-lived out-of-register states. To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event. We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences. We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance. Indeed, $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence. This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.
